---
title: "The validity of the multi-arm Big Stick Design"
author: "Diane Uschner"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("bsdRand.R"); source("getImbal.R")
```


## Background

The Big Stick Design has been shown to be highly effective to manage the 
trade-off between balance and randomness in two-arm clinical trials. 
It would therefore be desirable to generalize the design to multiple treatment arms.

Let $n$ be the total sample size and $K$ the number of treatment groups. 
Let $n_j(i)$ denote the number of patients in group $j$, $j = 1, \dots, K$, after the inclusion of patient 
$i$, $i=1,\dots, n$. Let the target
allocation ratio be $\frac{1}{K}$, resulting in a target group size $\frac{i}{K}$ after 
the allocation of patient $i$ in each of the groups $j \in \{1, \dots, K\}$. The difference 
of the actual group size to the target group size is thus $\tilde{D}_{ij} = n_j(i) - \frac{i}{k}$.
Restricting $\tilde{D}_{ij}$ is equivalent to restricting 
\[
D_{ij} = K\cdot n_j(i) - i.
\] 
For $k=2$, $D_{ij}$ reduces to the difference between treatment groups, 
as $D_{iA} = 2 \cdot n_A(i) - i = n_A(i) - n_B(i)$.

Let $b>0$ be the maximum tolerated Dalance. Generalizing the two-arm BSD, we want to assure 
$D_{ij} \leq b$ for all $i$ and $j$. 
Let $k_{i} = \#\{j: |D_{ij}| < b\}$ be the number of groups that are not at the imbalance boundary 
after the inclusion of patient $i$.
The following allocation rule has been proposed:
\[
P(T_{i+1} = j) = \begin{cases}
\frac{1}{k_i} & |D_{ij}| <b\\
0 & D_{ij} = b\\
1 & D_{ij} = -b
\end{cases}
\]

## Implementation

We want to provide an algorithm that reliably generates sequences within the 
imbalance boundary, and with the correct probability distribution. We implemented the 
following functions

* \texttt{bsdRand(N, b, K)} - Allocation algorithm for BSD with $N$ patients, $K$ treatment groups and imbalance boundary $b$.
* \texttt{getSum(R, K)} - For randomization sequence $R$, calculate cumulative number of patients in each of the $K$ treatment groups.
* \texttt{getImbal(R, K)} - For randomization sequence $R$, calculate cumulative imbalance in each of the $K$ treatment groups.
* \texttt{checkBsdSeq(R, b, K)} - Check whether R is a valid randomization sequence BSD(b, K).


### Three treatment groups

With three treatment groups the algorithm works fine. We see that the algorithm produces a valid BSD sequence that does not
exceed the imbalance boundary at any moment. At each moment only one group reaches the lower imbalance bound.

```{r}
N = 12;  K = 3; b = 2; set.seed(123)
(R <- bsdRand(N = N, b = b, K = K))
# is valid BSD(b) sequence?
checkBSDSeq(R, b, K)
# inspect n per treatment group after each allocation
getImbal(R, K)
# inspect treatment group imbalances
getImbal(R, K)
```

### Four treatment groups

But with more treatment groups there can occur problems:

```{r }
# results in stopping the algorithm early
N = 40;  K = 4; b = 6; set.seed(12412)
(R <- bsdRand(N = N, b = b, K = K))
```

Both groups 1 and 4 were at the lower imbalance boundary. Choosing either group would exceed the imbalance boundary for the other group in the next step.
We can look at the imbalance in each group to determine how it had come to that point:

```{r}
# Only look at realized allocations in R
R <- R[R != 0]
# is valid BSD(b) sequence?
checkBSDSeq(R, b, K)
# inspect n per treatment group after each allocation
getImbal(R, K)
# inspect treatment group imbalances
getImbal(R, K)
```

Clearly the imbalance exceed the maximum tolerated imbalance at various points during the process. The allocation breaks down because the more than two groups
are at the lower imbalance boundary, making a deterministic allocation impossible. 

## Algorithm 
The following algorithm implements the allocation rule presented above. Changes will be necessary to fulfill the criterion that the allocation sequences 
respect the maximum tolerated imbalance.

```{r}
N = 40; b = 6; K = 4; set.seed(12412)

R <- numeric(N) # Randomization sequence R = T
n <- rep(0, K) # Vector: n = (n_1(i), ...,n_K(i)) 
D <- rep(0, K) # Vector: D = (D_i1, ...,D_iK) 
i <- 0 # loop index

while(i < N) {
  # vector: atMti = (groups at -b, groups at b) 
  
  atMti <- list(which(D <= -b), which(D >= b))
  # vector: sumAtMti = (number of groups at -b, number of groups at b)
  sumAtMti <- sapply(atMti, length)
  
  # If no group at -b or b:  sample among all groups
  # INSTEAD: WILL NEED CRITERION ON NOT OVERSTEPPING THE BOUNDARY IN THE NEXT STEP
  if (sum(sumAtMti) == 0) {
    R[i + 1] <- sample(1:K, 1)
  # Deterministic allocation if one group is at -b
  } else if (sumAtMti[1] > 0){
  	# at each point, only one group can be at -b!
  	if (sumAtMti[1] > 1) {
  	  warning(paste("More than one group is at the lower imbalance boundary. Groups =", 
  	                  paste(atMti[[1]], collapse = ", "), ". Stopping early after", i, " allocations."))
  	  break 
  	}
    # Deterministic allocation if one group is at -b
  	R[i + 1] <- atMti[[1]]
  }
  # If some groups are at b: sample among those that are not at b
  else { # sumAtMti[2] > 0 
    notAtMti <- setdiff(1:K, atMti[[2]])
    # Deterministic allocation if there is only one group <b
    if(length(notAtMti) == 1){ 
    	R[i + 1] <- notAtMti
    }
    else{
    	R[i + 1] <- sample(notAtMti, 1)
    }
  } 
  i <- i + 1
  # increment the number of participants in the group that was selected
  n[R[i]] <- n[R[i]] + 1
  D <- K*n - i
}
R
```


